Unimodular covers of 3-dimensional parallelepipeds and Cayley sums

### 2019-07-13 unimodular-covers.tex
checking comments
several edits, mostly intro and Section 3

A lattice polytope is called \emph{smooth} if it is simple and the edge generators at every vertex form a linear basis for the lattice. 
%
Oda~\cite{Oda1997} posed several questions regarding smoothness and IDP for lattice polytopes. The following versions of them are now considered conjectures~\cite{HNPS2008,mfo2007}:
\begin{conjecture}
\label{conj:Oda}
\begin{enumerate}
\item 
\label{itm:smoothIDP}
(Related to problems 2 and 5 in \cite{Oda1997})
Every smooth lattice polytope is IDP.
\item 
\label{itm:mixedIDP}
(Related to problems 1, 3, 4, 6 in \cite{Oda1997}) Let $P, Q\subset \R^d$ be lattice polytopes with the normal fan of $P$ refining that of $Q$, and $Q$ smooth. Then
\begin{align*}
\label{eq:mixedIDP}
(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
\end{align*}
\end{enumerate}
\end{conjecture}

Motivated by  \Cref{conj:Oda}\eqref{itm:mixedIDP}, we recall the following definitions:

\begin{definition}
We say that $P$ is a \emph{weak Minkowski summand} of $Q$ if the normal fan of $Q$ refines that of $P$.
\end{definition}
The name is due to the fact that the condition on the refinement of the normal fans is equivalent to the existence of a polytope $P'$ such that $P+P' = k Q$ for some dilation constant $k>0$~\cite[??]{Grunbaum}. 
It has the following algebraic implication for the projective toric variety $X_Q$ corresponding to (the normal fan of) $Q$: $P$ is a weak Minkowski summand of $Q$ if and only if the corresponding Cartier divisor $D_P$ on  $X_Q$ is \emph{numerically effective}, or ``nef''~\cite[Cor.~6.2.15, Prop.~6.3.12]{CLS}; (observe that~\cite{CLS} use ``Minkowski summand'' meaning what we call ``weak Minkowski summand''). 
\paco{added this explanation}

\begin{definition}
The pair $(P, Q)$ has the integer decomposition property, or \emph{the pair $(P,Q)$ is IDP}, if 
\begin{align*}
(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
\end{align*}
\end{definition}

In this language,  \Cref{conj:Oda}\eqref{itm:mixedIDP} can be rephrased as: if $P$ is a weak Minkowski summand of $Q$ and $Q$ is smooth, then the pair $(P,Q)$ is IDP.

Motivated by these and other questions, several authors have studied the IDP property for different classes of lattice polytopes. 
%
For example,  very recently
Beck et al.~\cite{BHHHJKM2019} proved that all smooth centrally symmetric $3$-polytopes are IDP.
More precisely, they show that any such polytope can be covered by lattice 
parallelepipeds and unimodular simplices, both of which are trivially IDP.
%
In \Cref{sec:parallelepipeds} we show:

@@

Another case of Oda's questions that has been solved is the two-dimensional case of \Cref{conj:Oda}\eqref{itm:mixedIDP}, with three different proofs by Fakhruddin~\cite{Fakhruddin}, Ogata~\cite{Ogata} and Haase et al.~\cite{HNPS2008}, the latter without assuming $P$ smooth. 
In dimension three however the statement fails without the smoothness assumption, as shown by letting $P=Q$ be any non-unimodular empty tetrahedron.

An alternative approach to \Cref{conj:Oda}\eqref{itm:mixedIDP} is via Cayley sums, which we discuss in  \Cref{sec:cayley}. 
\paco{added def}
Recall that the \emph{Cayley sum} of two lattice polytopes $Q_1,Q_2\subset \R^d$ is the lattice polytope
\[
\cayley(Q_1,Q_2) := \conv(Q_1\times\{0\} \cup Q_2 \times \{1\}) \subset \R^3.
\]
\paco{new sentence. In fact, this implies that Cayley sums are the same as polytopes of width 1}
We normally require $\cayley(P,Q)$ to be full-dimensional (otherwise we can remove coordinates) but this does not need $Q_1$ and $Q_2$ to be full-dimensional. It only requires the linear subspaces parallel to them to span $\R^d$.

As we note there (\Cref{prop:mixedIDP}), if $P,Q\subset \R^d$ are two lattice polytopes one has
\[
\cayley(P,Q) \text{ is IDP} 
\Rightarrow
(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
\]
In particular, the following statement, proved in \Cref{sec:cayley}, is stronger than the afore-mentioned result of \cite{Fakhruddin,HNPS2008,Ogata}:

@@

This also implies the following two statements, also proved in \Cref{sec:cayley}.
A  \emph{prismatoid} is a polytope whose vertices all lie in two parallel facets. 
A polytope has width $1$ if its vertices lie in two \emph{consecutive} parallel lattice hyperplanes. Observe that this is the same as being $SL(\Z,d)$-equivalent to a Cayley sum.

@@

We believe that the $3$-polytopes in all these statements have unimodular triangulations, but this remains an open question.

@@

\begin{definition}
\label{def:circunpara}
Let $T$ be a tetrahedron with vertices $p_1$, $p_2$, $p_3$, and $p_4$. Consider the points $q_i= \frac12 (p_1+p_2+p_3+p_4) - p_i$, $i\ n [4]$, and let
\[
C(T)=\conv(p_i,q_i: i\in[4]).
\] 
$C(T)$ is a parallelepiped with facets $\conv(p_i, p_j, q_k, q_l)$ for all choices of $\{i,j,k,l\}=[4]$. We call it the \emph{parallelepiped circumscribed} to $T$.

@@

\paco{changed this}
The parallelepiped $P$ is the intersection of three regions of the form $f^{-1}([\alpha,\beta])$ for some functional $f\in (\R^3)^*$ and closed interval $[\alpha,\beta]\subset \R$. We will call such regions \emph{bands}. We claim that any band containing $T$ must contain at least three of the $q_i$s. This claim implies that the intersection $P$ of three bands contains at least one of the $q_i$s and hence $P$ contains the corresponding $T_i$.

@@

\begin{corollary}
\label{coro:coverpara}
Let $T$ be a empty lattice tetrahedron contained in a lattice parallelepiped $P$. Then, $T$ can be covered by unimodular tetrahedra contained in $P$.
\end{corollary}

\begin{proof}
We proceed by induction on the (normalized) volume of $T$, which is a positive integer. If this volume equals $1$ then $T$ is unimodular and there is nothing to prove, so we assume $T$ is not unimodular. As before, assume $T$ has vertices $p_1, p_2, p_3, p_4$.

\Cref{lemma:3<4} guarantees that one of the corner tetrahedra $T_i$ of the parallelepiped $C(T)$ is contained in $P$. Without loss of generality, suppose $T_4 = \conv(p_1, p_2, p_3,q_4)$ is in $P$. By \Cref{lemma:corner}, we know that $T_4$ contains a lattice point other than the $p_i$s, which we denote by $u$. 
%Observe that from the proof of \Cref{lemma:corner} we know more: $u$ lies in the face $\conv(p_1, p_2, q_4)$.\giulia{we need this for the volume computation below. perhaps it should be made explicit in the statemente of the lemma...}

Then $S=\conv(T\cup \{u\})$ can be triangulated in two different ways: $S=T \cup T'_4$, where $T'_4 = \conv(p_1, p_2, p_3, u) \subseteq T_4$ and $S= S_1 \cup S_2 \cup S_3$, with
\[
S_1= \conv(p_2,p_3,p_4, u),
S_2=\conv(p_1,p_3,p_4, u),
S_3=\conv(p_1,p_2,p_4, u).
\]
Each of the tetrahedra $S_i$ has lattice volume strictly smaller than $T$, because, for each $i$, $p_i$ is the unique point of $C(T)$ maximizing the distance to the opposite facet $\conv(p_j,p_k,p_l)$ of $T$. Thus, $S_1$, $S_2$ and $S_3$ cover $T$ and have volume strictly smaller than $T$. The $S_i$ may not be empty, but we can triangulate them into empty tetrahedra, which by inductive hypothesis they can be covered unimodularly, and we are done.

%Observe that  $T'_4$ and/or one of the $S_i$ can degenerate to a triangle, if $u$ lies in $\conv(p_1,p_2,p_3)$, but this does not affect the argument.
%
%: indeed, $\Vol(S_3)=\Vol(T'_4)$, since these tetrahedra have a common face $\conv(p_1, p_2, l)$ and the remaining vertex is at distance one from this face in both. Now $T'_4$ is contained in $T_4$, and we know that each corner tetrahedron $T_i$ has volume half that of $T$. The remaining two tetrahedra $S_1$ and $S_2$ must therefore have volume summing to that of $T$, and therefore each has volume strictly less than $T$.
\end{proof}

@@

We finish with a question which, if answered in the positive, would imply \Cref{conj:Oda}\eqref{itm:smoothIDP} in dimension three.
Let us say that a lattice $3$-polytope $P$ \emph{has the circumscribed parallelepiped property} if it satisfies the conclusion of \Cref{lemma:3<4}:  ``for every empty tetrahedron $T$ contained in $P$ at least one of the four corner tetrahedra in $C(T)$ is 
contained in $P$''. 
If this holds then $P$ has a unimodular cover, since then the proofs of \Cref{coro:coverpara} and \Cref{thm:parallelepipeds} work for $P$.

\begin{question}
Does every smooth 3-polytope have the circumscribed parallelepiped property? By the above remark, if the answer is yes then every smooth $3$-polytope has a unimodular cover.
\end{question}

@@

\section{ Cayley sums} \giulia{Change notations: remove nef and make sure everything works}
\label{sec:cayley}
Let $P$ and $Q$ be two lattice polytopes in $\R^d$. We do not require them to be full-dimensional, but we assume their Minkowski sum is. Remember that the Minkowski sum $P+Q$ of $P$ and $Q$ is defined as:
\[
P + Q := \{ p+q \in \R^d: p\in P, q \in Q\}.
\]
The \emph{Cayley sum} of $P$ and $Q$ is defined as
\[
\cayley (P,Q) = \conv( P\times\{0\} \cup Q\times \{1\}) \in \R^{d+1}.
\]
The assumption that $P+Q$ is $d$-dimensional is equivalent to $\cayley (P,Q)$ being $(d+1)$-dimensional.

\begin{definition}
In the above conditions, we say that $Q$ \emph{is a weak Minkowski summand of $P$} if the normal fan of $Q$ refines that of $P$. 
If this happens, we say that $\cayley(P,Q)$ is a \emph{nef Cayley sum}.\giulia{I guess there's no nice term for this now...}
\paco{not sure; maybe we can keep nef after all; another (unclear) option could be ``weak prism over $Q$'' or ``degenerate prism over $Q$''.}

We say that the pair $(P,Q)$ is \emph{mixed IDP} if 
\paco{``mixed IDP'' is not the right property, since this is weaker than IDP} \giulia{do we give it a name at all then?}
\paco{perhaps mixed-IDP}
\[
(P\cap \Z^d) + (Q\cap \Z^d) = (P+Q)\cap \Z^d.
\]
\end{definition}

In this language, \Cref{conj:Oda}\eqref{itm:mixedIDP} translates to:
\emph{If $Q$ is nef for $P$ and $P$ is smooth then the pair $(P,Q)$ mixed IDP.}

\begin{proposition}
\label{prop:mixedIDP}
If $\cayley(P,Q)$ is IDP then the pair $(P,Q)$ is mixed IDP.
\end{proposition}

The so-called \emph{Cayley Trick} is the isomorphism
\[
2\cayley(P, Q) \cap (\R^d\times \{1\}) \cong P+Q,
\]
which induces the following canonical bijections:

\begin{center}
  \begin{tabular}{ c | c }
    $\cayley(P,Q)$ & $P+Q$\\
    \hline
   polyhedral subdivisions & mixed subdivisions \\
    triangulations & fine mixed subdivisions \\
   unimodular simplices & unimodular prod-simplices \\
  \end{tabular}
\end{center}\giulia{not sure if the table makes it clear that these are bijections}

\begin{align*}
\text{polyhedral subdivisions of $\cayley(P,Q)$} &\leftrightarrow \text{mixed subdivisions of $P + Q$}\\
\text{triangulations of $\cayley(P,Q)$} &\leftrightarrow \text{fine mixed subdivisions of $P + Q$}\\
\text{unimodular simplices in $\cayley(P,Q)$} &\leftrightarrow \text{unimodular prod-simplices in $P + Q$}.
\end{align*}
See \cite{DLRS2010} for more details on the Cayley Trick and on triangulations and polyhedral subdivisions of polytopes.
In fact these bijections can be taken as definitions of the objects in the right-hand sides. In particular, 
we call \emph{prod-simplices} in $P+Q$ the Minkowski sums $T_1+T_2$ where $T_1\subset P$ and $T_2\subset Q$ are simplices with complementary affine spans. A prod-simplex is \emph{unimodular} if the edge vectors from a vertex of $T_1$ and from a vertex of $T_2$ form a unimodular basis.

@@

We now turn our attention to dimension $d=2$.  The unimodular prod-simplices are unimodular parallelograms $p+q$ with $p$ a segment of $P$ and $q$ a segment of $Q$, and unimodular triangles $T+u$, where $T \subset P$ and $u \in Q$ or viceversa.

Consider any lattice triangulation $\T$ of $\cayley(P,Q)$, or equivalently a mixed subdivision of $P+Q$. Observe that any simplex in $T$ has vertices in $P \times \{0\}$ or in $Q \times \{1\}$, since $\cayley(P,Q)$ lies between two consecutive lattice hyperplanes $\{x_3=0\}$ and $\{x_3=1\}$. Thus simplices in $\T$ are of three types; they can , or have three vertices in $P$ and one in $Q$ (type $(3,1)$), have one vertex in $P$ and three in $Q$ (type $(1,3)$) or have two vertices in $P$ and two in $Q$ (type $(2,2)$).

\giulia{fix this}

%As a consequence, we also have that:
%\begin{align*}
%\text{$\cayley(P,Q)$ has a unimodular cover} &\leftrightarrow \text{$P+Q$ can be covered by unimodular prod-simplices}.
%\end{align*}

Our proof of \Cref{thm:cayley} is based on the following result, the proof of which we postpone to \Cref{sec:the_lemma}. 
The lemma uses the following conventions. If $a, b$ are points, we denote by $[a,b]$ and $(a,b)$ respectively the closed and open segments with endpoints $a,b$. For a segment $s=[a,b]$, we denote the vector $\vec s:= b-a$ and $\vecline s$ as the line spanned by $\vec s$.

\begin{lemma}
\label{lemma:cayley}
Let $P$ and $Q$ be two lattice polygons with $Q$ nef with respect to $P$. Let $p=[p_1,p_2] \subset P$ and $q=[q_1,q_2]\subset Q$ be two primitive and non-parallel lattice segments, and let $\vecline p$ and $ \vecline q$ be the lines spanned by them.  If the parallelogram $p + q$ is not unimodular, then at least one of the regions
\[
((p_1, p_2) + \vecline q ) \cap P, 
\qquad \text{and} \qquad
((q_1, q_2) + \vecline p ) \cap Q
\]
contains a lattice point. See \Cref{fig:strips}.
%\[
%(\int(p) + \vecline q) \cap P
%\qquad \text{or} \qquad
%\int(q + \vecline p) \cap Q
%\]
%contains a lattice point.
\end{lemma}
\giulia{what I want is: a lattice point $p \in P$ such that either $p+q_1$ or $p+q_2$ lies in the interior of $\{p_1, p_2\} + \{q_1, q_2\}$ (or symmetric for $Q$)}

@@

\begin{proof}[Proof of \Cref{thm:cayley}]

A unimodular triangulation of a simplex of type $(3,1)$ (resp $(1,3)$) is obtained simply by unimodularly triangulating the triangular face lying in $P$ (resp $Q$) and coning over the vertex in $Q$ (resp $P$). This leaves simplices of type $(2,2)$. In the mixed subdivision of $P+Q$ corresponding to $T$, these are mixed cells, that is parallelograms with one edge in $P$ and one in $Q$ \giulia{how imprecise can be without being confusing?}.

Let $S=\{p_1, p_2\} + \{q_1, q_2\}$ be such a parallelogram. By \Cref{lemma:cayley} we know that there is a lattice point $p \in P$ such that either $p+q_1$ or $p+q_2$ lies in $S$, w.l.o.g. suppose it is $p+q_1$. Then we can cover $S$ by the triangle $\{p_1, p_2, p\} + q_1$ and parallelograms $\{p_1, p\}  + \{q_1, q_2\}$ and $\{p_2, p\} + \{q_1, q_2\}$. As above, the triangle can be unimodularly triangulated, while the volume of the two parallelograms sums up to the volume of $S$, and therefore each has volume strictly smaller than $S$. By iterating this procedure, we can cover $S$ by unimodular cells.
\end{proof}

We now turn our attention to \emph{prismatoids}, defined in~\cite{Santos-hirsch} as polytopes whose vertices all lie in two parallel facets. In particular, a \emph{lattice prismatoid} is any $d$-polytope $SL(\Z,d)$-equivalent to one of the form
\[
\conv(Q_1\times\{0\} \cup Q_2 \times \{k\}),
\]
where $Q_1,Q_2$ are lattice $(d-1)$-polytopes and $k\in \Z_{>0}$. This generalizes Cayley sums, which are the case $k=1$.

>>>>

Oda~\cite{Oda1997} posed several questions regarding smoothness and the IDP property for lattice polytopes.
%
\paco{I think this definition goes better here, and decided to "undeclare" it and the definition of weak Mink sum}
Following \cite{HaaseHof, Tsuchiya} we say that a pair $(P, Q)$ of lattice polytopes has the integer decomposition property, or that \emph{the pair $(P,Q)$ is IDP}, if 
\begin{align*}
\label{eq:mixedIDP}
(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
\end{align*}
%
A lattice polytope $Q$ is called \emph{smooth} if it is simple and the edge generators at every vertex form a linear basis for the lattice; equivalently, if the projective toric variety defined by the normal fan of $Q$ is smooth. 
The following versions of Oda's questions are now considered conjectures~\cite{HNPS2008,mfo2007}, and they are open even in dimension three:
\begin{conjecture}
\label{conj:Oda}
\begin{enumerate}
\item 
\label{itm:smoothIDP}
(Related to problems 2 and 5 in \cite{Oda1997})
Every smooth lattice polytope is IDP.
\item 
\label{itm:mixedIDP}
(Related to problems 1, 3, 4, 6 in \cite{Oda1997}) Every pair $(P,Q)$ of lattice polytopes with $Q$ smooth and the normal fan of $Q$ refining that of $P$ is IDP.
%\begin{align}
%\label{eq:mixedIDP}
%(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
%\end{align}
\end{enumerate}
\end{conjecture}

When the normal fan of a $Q$ refines that of $P$, as in the second conjecture, we say that $P$ \emph{is a weak Minkowski summand of $Q$}, since this is equivalent to the existence of a polytope $P'$ such that $P+P' = k Q$ for some dilation constant $k>0$\paco{did not find this in the google books version of \cite{Grunbaum}; found it in Postnikov-Reiner-Williams, Tsuchiya, a survey by Rekha Thomas, and other places. Perhaps we can leave it wothout reference}. 
This property has the following algebraic implication for the projective toric variety $X_Q$: $P$ is a weak Minkowski summand of $Q$ if and only if the Cartier divisor on  $X_Q$ defined by $P$ is \emph{numerically effective}, or ``nef'' (see~\cite[Cor.~6.2.15, Prop.~6.3.12]{CLS}, but observe that there ``Minkowski summand'' is used meaning what we call ``weak Minkowski summand''). 
%\paco{added this explanation}
%
%\begin{definition}
%Following \cite{HaaseHof, Tsuchiya} we say that the pair $(P, Q)$ has the integer decomposition property, or \emph{the pair $(P,Q)$ is IDP}, if 
%\begin{align*}
%(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
%\end{align*}
%\end{definition}
% 
%In this language,  \Cref{conj:Oda}\eqref{itm:mixedIDP} can be rephrased as: if $P$ is a weak Minkowski summand of a smooth polytope $Q$ then the pair $(P,Q)$ is IDP.

\medskip 
Motivated by these and other questions, several authors have studied the IDP property for different classes of lattice polytopes. 
%
For example,  very recently
Beck et al.~\cite{BHHHJKM2019} proved that all smooth centrally symmetric $3$-polytopes are IDP.
More precisely, they show that any such polytope can be covered by lattice 
parallelepipeds and unimodular simplices, both of which are trivially IDP.
%
In \Cref{sec:parallelepipeds} we show:

@@

The two-dimensional case of \Cref{conj:Oda}\eqref{itm:mixedIDP} is known to hold, with three different proofs by Fakhruddin~\cite{Fakhruddin}, Ogata~\cite{Ogata} and Haase et al.~\cite{HNPS2008}. This last one actually shows that smoothness of $Q$ is not needed. In dimension three, however, the conjecture fails without the smoothness assumption, as shown by letting $P=Q$ be any non-unimodular \emph{empty tetrahedron}, by which we mean a lattice tetrahedron with no lattice points other than its vertices (see the proof of \Cref{lemma:corner} for details on them).
\paco{added def of empty tetrahedron}

An alternative approach to \Cref{conj:Oda}\eqref{itm:mixedIDP} is via Cayley sums, which we discuss in  \Cref{sec:cayley}. 
Recall that the \emph{Cayley sum} of two lattice polytopes $Q_1,Q_2\subset \R^d$ is the lattice polytope
\[
\cayley(Q_1,Q_2) := \conv(Q_1\times\{0\} \cup Q_2 \times \{1\}) \subset \R^3.
\]
We normally require $\cayley(P,Q)$ to be full-dimensional (otherwise we can delete coordinates) but for this does not need $Q_1$ and $Q_2$ to be full-dimensional. It only requires the linear subspaces parallel to them to span $\R^d$.

As we note in \Cref{prop:mixedIDP}, if the Cayley sum of $P$ and $Q$ is IDP then the pair $(P,Q)$ is IDP.
In particular, the following statement from \Cref{sec:cayley} is stronger than the afore-mentioned result of \cite{Fakhruddin,HNPS2008,Ogata}:

@@

This has the following two consequences, also proved in \Cref{sec:cayley}.
A  \emph{prismatoid} is a polytope whose vertices all lie in two parallel facets. 
A polytope has width $1$ if its vertices lie in two \emph{consecutive} parallel lattice hyperplanes. Observe that this is the same as being ($SL(\Z,d)$-equivalent to) a Cayley sum.

@@

\medskip
We believe that the $3$-polytopes in all these statements have unimodular triangulations, but this remains an open question.

@@

\begin{definition}
\label{def:circunpara}
Let $T$ be a tetrahedron with vertices $p_1$, $p_2$, $p_3$, and $p_4$. Consider the points $q_i= \frac12 (p_1+p_2+p_3+p_4) - p_i$, $i\in [4]$, and let
\[
C(T)=\conv(p_i,q_i: i\in[4]).
\] 
$C(T)$ is a parallelepiped with facets $\conv(p_i, p_j, q_k, q_l)$ for all choices of $\{i,j,k,l\}=[4]$. We call it the \emph{parallelepiped circumscribed} to $T$.

@@

\paco{changed this}
We call \emph{band} any region of the form $f^{-1}([\alpha,\beta])$ for some functional $f\in (\R^3)^*$ and closed interval $[\alpha,\beta]\subset \R$.
We claim that any band containing $T$ must contain at least three of the $q_i$s. 
This claim implies that the parallelepiped $P$, which is the intersection of three bands, contains at least one of the $q_i$s and hence it fully contains the corresponding $T_i$.

@@

\begin{corollary}
\label{coro:coverpara}
Let $T$ be an empty lattice tetrahedron contained in a lattice parallelepiped $P$. Then, $T$ can be covered by unimodular tetrahedra contained in $P$.
\end{corollary}

\begin{proof}
We proceed by induction on the (normalized) volume of $T$, which is a positive integer. If this volume equals $1$ then $T$ is unimodular and there is nothing to prove, so we assume $T$ is not unimodular. Let $p_1, p_2, p_3, p_4$ denote the vertices of $T$.

\Cref{lemma:3<4} guarantees that one of the corner tetrahedra $T_i$ of the parallelepiped $C(T)$ is contained in $P$. Without loss of generality, suppose $T_4 = \conv(p_1, p_2, p_3,q_4)$ is in $P$. By \Cref{lemma:corner}, we know that $T_4$ contains a lattice point other than the $p_i$s, which we denote by $u$. 
%Observe that from the proof of \Cref{lemma:corner} we know more: $u$ lies in the face $\conv(p_1, p_2, q_4)$.\giulia{we need this for the volume computation below. perhaps it should be made explicit in the statemente of the lemma...}
%
Then $S=\conv(T\cup \{u\})$ can be triangulated in two different ways: $S=T \cup T'_4$, where $T'_4 = \conv(p_1, p_2, p_3, u) \subseteq T_4$ and $S= S_1 \cup S_2 \cup S_3$, with
\[
S_1= \conv(p_2,p_3,p_4, u),
S_2=\conv(p_1,p_3,p_4, u),
S_3=\conv(p_1,p_2,p_4, u).
\]

Each of the tetrahedra $S_i$ has lattice volume strictly smaller than $T$ because, for each $i$, $p_i$ is the unique point of $C(T)$ maximizing the distance to the opposite facet $\conv(p_j,p_k,p_l)$ of $T$. Thus, $S_1$, $S_2$ and $S_3$ cover $T$ and have volume strictly smaller than $T$. The $S_i$ may not be empty, but we can triangulate them into empty tetrahedra, which by inductive hypothesis they can be covered unimodularly.
%
%Observe that  $T'_4$ and/or one of the $S_i$ can degenerate to a triangle, if $u$ lies in $\conv(p_1,p_2,p_3)$, but this does not affect the argument.
%
%: indeed, $\Vol(S_3)=\Vol(T'_4)$, since these tetrahedra have a common face $\conv(p_1, p_2, l)$ and the remaining vertex is at distance one from this face in both. Now $T'_4$ is contained in $T_4$, and we know that each corner tetrahedron $T_i$ has volume half that of $T$. The remaining two tetrahedra $S_1$ and $S_2$ must therefore have volume summing to that of $T$, and therefore each has volume strictly less than $T$.
\end{proof}

@@

%We finish with a question which, if answered in the positive, would imply \Cref{conj:Oda}\eqref{itm:smoothIDP} in dimension three.
Let us say that a lattice $3$-polytope $P$ \emph{has the circumscribed parallelepiped property} if it satisfies the conclusion of \Cref{lemma:3<4}:  ``for every empty tetrahedron $T$ contained in $P$ at least one of the four corner tetrahedra in $C(T)$ is 
contained in $P$''. 
If this holds then $P$ has a unimodular cover, since then the proofs of \Cref{coro:coverpara} and \Cref{thm:parallelepipeds} work for $P$.
Hence, a positive answer to the following question would imply that every smooth $3$-polytope has a unimodular cover, which in turn implies \Cref{conj:Oda}\eqref{itm:smoothIDP} in dimension three.

\begin{question}
Does every smooth 3-polytope have the circumscribed parallelepiped property? 
\end{question}

@@

\section{ Cayley sums} \giulia{Change notations: remove nef and make sure everything works}
\label{sec:cayley}
Let $P$ and $Q$ be two lattice polytopes in $\R^d$. We do not require them to be full-dimensional, but we assume their Minkowski sum is. Remember that the \emph{Minkowski sum} $P+Q$ and the \emph{Cayley sum} of $P$ and $Q$ are defined as:
\begin{gather*}
P + Q := \{ p+q \in \R^d: p\in P, q \in Q\}.\\
\cayley (P,Q) = \conv( P\times\{0\} \cup Q\times \{1\}) \in \R^{d+1}.
\end{gather*}

\paco{removed definitions of "nef cayley", "mixed IDP" and "weak Minkowski sum" from here}
%\begin{definition}
%In the above conditions, we say that $Q$ \emph{is a weak Minkowski summand of $P$} if the normal fan of $Q$ refines that of $P$. 
%If this happens, we say that $\cayley(P,Q)$ is a \emph{nef Cayley sum}.\giulia{I guess there's no nice term for this now...}
%\paco{not sure; maybe we can keep nef after all; another (unclear) option could be ``weak prism over $Q$'' or ``degenerate prism over $Q$''.}
%
%We say that the pair $(P,Q)$ is \emph{mixed IDP} if 
%\paco{``mixed IDP'' is not the right property, since this is weaker than IDP} \giulia{do we give it a name at all then?}
%\paco{perhaps mixed-IDP}
%\[
%(P\cap \Z^d) + (Q\cap \Z^d) = (P+Q)\cap \Z^d.
%\]
%\end{definition}
%
%In this language, \Cref{conj:Oda}\eqref{itm:mixedIDP} translates to:
%\emph{If $Q$ is nef for $P$ and $P$ is smooth then the pair $(P,Q)$ mixed IDP.}
%
The so-called \emph{Cayley Trick} is the isomorphism
\[
2\cayley(P, Q) \cap (\R^d\times \{1\}) \cong P+Q,
\]
which easily implies:

\begin{proposition}[see, e.g.~\protect{\cite[Thm.~0.4]{Tsuchiya}}]
\label{prop:mixedIDP}
If $\cayley(P,Q)$ is IDP then the pair $(P,Q)$ is mixed IDP.
\end{proposition}

%The assumption that $P+Q$ is $d$-dimensional is equivalent to $\cayley (P,Q)$ being $(d+1)$-dimensional.

The Cayley Trick also provides the following canonical bijections:

\begin{center}
  \begin{tabular}{ c c c }
    $\cayley(P,Q)$ && $P+Q$\\
    \hline
   polyhedral subdivisions &$\leftrightarrow$& mixed subdivisions \\
    triangulations &$\leftrightarrow$& fine mixed subdivisions \\
   unimodular simplices &$\leftrightarrow$& unimodular prod-simplices \\
  \end{tabular}
\end{center}\giulia{not sure if the table makes it clear that these are bijections}

\[
\begin{array}{ccc}
\text{polyhedral subdivisions of $\cayley(P,Q)$} &\leftrightarrow& \text{mixed subdivisions of $P + Q$}\\
\text{triangulations of $\cayley(P,Q)$} &\leftrightarrow& \text{fine mixed subdivisions of $P + Q$}\\
\text{unimodular simplices in $\cayley(P,Q)$} &\leftrightarrow& \text{unimodular prod-simplices in $P + Q$}.
\end{array}
\]
See \cite{DLRS2010} for more details on the Cayley Trick and on triangulations and polyhedral subdivisions of polytopes.
In fact these bijections can be taken as definitions of the objects in the right-hand sides. In particular, 
we call \emph{prod-simplices} in $P+Q$ the Minkowski sums $T_1+T_2$ where $T_1\subset P$ and $T_2\subset Q$ are simplices with complementary affine spans. A prod-simplex is \emph{unimodular} if the edge vectors from a vertex of $T_1$ and from a vertex of $T_2$ form a unimodular basis.

@@

We now turn our attention to $d=2$, in order to prove \Cref{thm:cayley}. In a triangulation of $\cayley(P,Q)\subset \R^3$ there are tetrahedra of types $(1,3)$, $(2,2)$ and $(3,1)$, where the type denotes how many vertices they have in $P$ and in $Q$. Empty tetrahedra of types $(1,3)$ or $(3,1)$ are automatically unimodular, so we only need to care with those of type $(2,2)$. These are the Cayley sums of a segment $p\subset P$ and a segment $q\subset Q$.
%
%
%The unimodular prod-simplices are unimodular parallelograms $p+q$ with $p$ a segment of $P$ and $q$ a segment of $Q$, and unimodular triangles $T+u$, where $T \subset P$ and $u \in Q$ or viceversa. 
%
%Consider any lattice triangulation $\T$ of $\cayley(P,Q)$, or equivalently a mixed subdivision of $P+Q$. Observe that any simplex in $T$ has vertices in $P \times \{0\}$ or in $Q \times \{1\}$, since $\cayley(P,Q)$ lies between two consecutive lattice hyperplanes $\{x_3=0\}$ and $\{x_3=1\}$. Thus simplices in $\T$ are of three types; they can , or have three vertices in $P$ and one in $Q$ (type $(3,1)$), have one vertex in $P$ and three in $Q$ (type $(1,3)$) or have two vertices in $P$ and two in $Q$ (type $(2,2)$). 
%
\giulia{fix this}
\paco{done, but check it}
%
%As a consequence, we also have that:
%\begin{align*}
%\text{$\cayley(P,Q)$ has a unimodular cover} &\leftrightarrow \text{$P+Q$ can be covered by unimodular prod-simplices}.
%\end{align*}
For this the following result, the proof of which we postpone to \Cref{sec:the_lemma}, is crucial. 
The lemma uses the following conventions. If $a, b$ are points, we denote by $[a,b]$ and $(a,b)$ respectively the closed and open segments with endpoints $a,b$. For a segment $s=[a,b]$, we denote the vector $\vec s:= b-a$ and $\vecline s$ as the line spanned by $\vec s$.

\begin{lemma}
\label{lemma:cayley}
\paco{$Q$ is 2-dim, but $P$ may not}
Let $Q$ be a two-dimensional lattice polytope and $P$ a weak Minkowski summand of it. 
Let $p=[p_1,p_2] \subset P$ and $q=[q_1,q_2]\subset Q$ be two primitive and non-parallel lattice segments, and let $\vecline p$ and $ \vecline q$ be the lines spanned by them.  If the parallelogram $p + q$ is not unimodular, then at least one of the regions
\[
((p_1, p_2) + \vecline q ) \cap P, 
\qquad \text{and} \qquad
((q_1, q_2) + \vecline p ) \cap Q
\]
contains a lattice point. See \Cref{fig:strips}.
%\[
%(\int(p) + \vecline q) \cap P
%\qquad \text{or} \qquad
%\int(q + \vecline p) \cap Q
%\]
%contains a lattice point.
\end{lemma}
%\giulia{what I want is: a lattice point $p \in P$ such that either $p+q_1$ or $p+q_2$ lies in the interior of $\{p_1, p_2\} + \{q_1, q_2\}$ (or symmetric for $Q$)}

@@

\begin{corollary}
\label{coro:covercayley}
\paco{inserted this corollary to make proof clearer, and to parallel section 2}
Let $T$ be an empty lattice tetrahedron contained in the Cayley sum $\cayley(P,Q)$, where $P$ is a weak Minkowski summand of $Q$. Then, $T$ can be covered by unimodular tetrahedra contained in $\cayley(P,Q)$.
\end{corollary}

\begin{proof}
The proof is by induction on the normalized volume of $T$, which we assume to be at least $2$. This implies that $T$ is of type $(2,2)$,
since empty tetrahedra of types $(1,3)$ and $(3,1)$ are unimodular. Thus, $T$ is the Cayley sum of primitive segments $p=[p_1,p_2]\subset P$ and $q=[q_1,q_2]\subset Q$.  
Let $u$ be the lattice point whose existence is guaranteed by \Cref{lemma:cayley}. Assume  (the other case is similar) that 
\[
u \in ((p_1, p_2) + \vecline q ) \cap P,
\]
and call $t$ the triangle $t=\conv( u,  p_1,  p_2)\subset P$.
